Lossy compression of DNA sequencing quality data

### 2015-04-23 manuscript/q-compression.tex
Numerous .small edits

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ --- profiling
  with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponds to original uncompressed
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

As the order of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values. A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per sequence of quality values. The read length of the \textit{Mus musculus}
data set is only 26, so the 7th-degree polynomial regression is storing
six more bytes than necessary for lossless encoding the quality data.

@@

It is important to highlight that the even though a compression method
may result in the same number of filtered base-pairs as the
uncompressed sequences, it does not mean the \emph{same} base-pairs
were filtered. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other. However, if we examine the specific
reads filtered, the 5th-degree polynomial regression discards
approximately two thirds less reads than the 1st-degree polynomial
regression that are kept by the uncompressed reads (4,640 and
12,066 reads, respectively) (Supplemental Table 2).

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
outperforms all others in all metrics (Table
\ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
%using the uncompressed reads outperforms all compression methods in
%terms of \textsc{lap}, NG50, relocations, translations, and
%inversions.
Among the compression methods, the 256-, 64-, 128-profile encoding had
the highest ranks, followed by \textsc{q}ual\textsc{c}omp with rates 10 bits/read, 30
bits/read, 100 bits/read, then 7th-degree polynomial regression,
followed by the \textsc{q}ual\textsc{c}omp with rate 6 bits/read, the 2-bin encoding,
and lastly, the 3rd-degree, 5th-degree, and 0-degree polynomials.

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see
Supplemental Table 3). Despite the similar contiguity statistics,
the different compression methods vary noticeably in the amount of
\textsc{snp}s. The order of polynomial has an inverse relationship
with the amount of \textsc{snp}s detected. The 2-bin and profile
methods detected the least amount of \textsc{snp}s compared to the
reference genome, outperforming the assembly using the original
quality values. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual \textsc{snp}s.

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads scores worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using uncompressed reads has an error rate of roughly 8.75
errors per 100 kb of assembled sequence, while the 256-profile
encoding has an error rate of 8.02 errors per 100 kbp (Supplemental
Table 3).

In general, the greater the polynomial order, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poor in all metrics except assembled bases.
One explanation is that the high error portions of reads are being
marked as high quality, so \textsc{allpaths-lg} is unable to trim or
error correct the sequences. Assembled sequences that overlap maybe
unable to align across the errors at the end of the reads, artificially
inflating the assembled genome size.

@@

\begin{figure}[!tbp]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}

@@

A potential extension to 2-bin encoding is to incorporate an
additional bin (\emph{okay}). The \emph{okay} value can be used where
the base qualities fall within a 2-bin range. Because the distribution
of quality values is skewed towards higher quality, we need to
experiment with different cutoffs for the \emph{okay} value and
determine if the additional storage is noticeable in subsequent
analyses.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality sequences using two
high-coverage bacterial, human, and mouse data sets.
A more direct comparison against additional lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
performed. Furthermore, other types of sequencing data can be
analyzed apart from the Illumina data examined here. For example, the
PacBio sequencing instruments produce very long reads (with average
read lengths on the order of 10 kbp), but with the trade-off of having
a high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

@@

\section*{Acknowledgement}
We thank the 2014 Bioinformatics
Exchange for Students and Teachers (\textsc{best}) Summer School, 
funded by the offices of the Dean of The Graduate School at University
of Maryland and the Rektor of University of T\"{u}bingen, where this 
research was initiated.

>>>>

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponding to original
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values. A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per sequence of quality values. The read length of the \textit{Mus
  musculus} data set is only 26, so the 7th-degree polynomial
regression is storing six more bytes than necessary for lossless
encoding the quality data.

@@

It is important to highlight that the even though a compression method
may result in the same number of filtered base calls as the
uncompressed sequences, it does not mean the \emph{same} base calls
were filtered. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other. However, if we examine the specific reads
filtered, the 5th-degree polynomial regression discards approximately
two thirds less reads than the 1st-degree polynomial regression that
are kept by the uncompressed reads (4,640 and 12,066 reads,
respectively) (Supplemental Table 2).

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
outperforms all others in all metrics (Table
\ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
%using the uncompressed reads outperforms all compression methods in
%terms of \textsc{lap}, NG50, relocations, translations, and
%inversions.
Among the compression methods, the profile encoding performed best,
polynomial regression modeling with degree $>$ 7 performed worst, and
other methods were intermediate (Fig.~\ref{fig:assembly_ranks}).

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbp (see
Supplemental Table 3). Despite the similar contiguity statistics, the
different compression methods vary noticeably in the amount of
\textsc{snp}s. The 2-bin and profile methods exhibited the fewest
\textsc{snp}s compared to the reference genome, thus outperforming the
assembly using the original quality values in this characteristic. A
more in-depth evaluation is needed to determine whether these
compression methods are missing actual \textsc{snp}s.

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads scores worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using the original uncompressed reads has an error rate of
roughly 8.75 errors per 100 kb of assembled sequence, while the
256-profile encoding has an error rate of 8.02 errors per 100 kbp
(Supplemental Table 3).

In general, the greater the polynomial degree, the better overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poorly in all metrics except assembled
bases.  One explanation is that the high error portions of reads are
being marked as high quality, so \textsc{allpaths-lg} is unable to
trim or error correct the sequences. Assembled sequences that overlap
maybe unable to align across the errors at the end of the reads,
artificially inflating the assembled genome size.

@@

\begin{figure}[!tbp]
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}

@@

A potential extension to 2-bin encoding is to incorporate an
additional bin (\emph{okay}). The \emph{okay} value can be used where
the base qualities fall within a 2-bin range. Because the distribution
of quality values is skewed towards higher quality, we need to
experiment with different cutoffs for the \emph{okay} value and
determine if the additional storage costs are associated with positive
effects on subsequent analyses.

@@

We have simply provided here the initial steps in analyzing the effect
of lossy compression on quality values using high-coverage bacterial,
human, and mouse data sets.  A more direct comparison against
additional lossy compression tools, such as
\textsc{SlimGene}~\citep{Kozanitis:2011kl}, needs to be
performed. Furthermore, other types of sequencing data can be analyzed
apart from the Illumina data examined here. For example, the PacBio
sequencing instruments produce very long reads (with average read
lengths on the order of 10 kbp), but with the trade-off of having a
high error-rate ($\sim$15\%). Unlike the class of quality values we
have examined here, the distribution of erroneous bases is relatively
uniform~\cite{Ferrarini:2013vf}. The assembly complexity of bacterial
genomes can be greatly simplified, producing near complete genome
assemblies, by utilizing a single run of these long
reads~\citep{Koren:2013ye}. If long read sequencing technologies such
as PacBio become more widely adopted, it would be of huge benefit to
examine the potential of lossy compression algorithms on not only the
quality values, but the biological sequencing data themselves.

@@

\section*{Acknowledgement}
This research was initiated at the 2014 Bioinformatics Exchange for
Students and Teachers (\textsc{best}) Summer School, which was funded
by the offices of the Dean of The Graduate School at University of
Maryland and the Rektor of University of T\"{u}bingen.
